---
title: "LST GAM Analysis - Response-Specific & LCZ-Stratified"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(sf)
library(dplyr)
library(ggplot2)
library(mgcv)
library(tidyr)
library(GGally)
library(car)
library(purrr)
library(broom)
library(tibble)
```

---

## 1. Load Data

```{r load-data}
INPUT_GPKG <- "../04_LCZ_Classifications/02_LCZ_Classification/LCZ_Classification.gpkg"

raw_gdf <- st_read(INPUT_GPKG, quiet = TRUE)
print(names(raw_gdf))
```


---

## 2. Sanity Check

```{r sanity-check}
period_cols <- intersect(c("Night", "Morning", "Afternoon", "Evening"), names(raw_gdf))
cat("Period columns found:", paste(period_cols, collapse = ", "), "\n")

cat("\n=== Missingness ===\n")
raw_gdf %>%
  st_drop_geometry() %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  print()
```

---

## 3. Build Dataset (wide format, four response variables)

```{r build-dataset}
lst_wide <- raw_gdf %>%
  select(
    GRID_ID,
    all_of(period_cols),
    Tree_Canopy = TreeCanopy_mean,
    Impervious = ImperviousSurface_mean,
    Bldg_Footprint = BuildingFootprint_mean,
    LCZ = LCZ_final
  ) %>%
  mutate(
    Average = rowMeans(st_drop_geometry(.)[period_cols], na.rm = TRUE),
    LCZ = as.factor(LCZ)
  )

RESPONSE_VARS <- c(period_cols, "Average")
cat("Response variables:", paste(RESPONSE_VARS, collapse = ", "), "\n")

# rows usable for predictor/LCZ analysis (each response's own NA rows
# get dropped separately at model-fit time, not here)
lst_predictors_complete <- lst_wide %>%
  drop_na(Tree_Canopy, Impervious, Bldg_Footprint, LCZ)

cat("\nRows with complete predictors + LCZ:", nrow(lst_predictors_complete),
    "/", nrow(lst_wide), "\n")

cat("\n=== LCZ Distribution ===\n")
table(lst_predictors_complete$LCZ)
```

---

## 4. Redundancy Screening

### 4.1 Correlation Matrix (predictors + all four responses)

```{r correlation-matrix}
cor_mat <- lst_wide %>%
  st_drop_geometry() %>%
  select(all_of(RESPONSE_VARS), Tree_Canopy, Impervious, Bldg_Footprint) %>%
  cor(use = "complete.obs")

round(cor_mat, 2)

GGally::ggcorr(
  lst_wide %>%
    st_drop_geometry() %>%
    select(all_of(RESPONSE_VARS), Tree_Canopy, Impervious, Bldg_Footprint),
  label = TRUE
)
```

### 4.2 VIF (predictors only — no response, no LCZ)

```{r vif-check}
lm_vif <- lm(Average ~ Tree_Canopy + Impervious + Bldg_Footprint,
             data = lst_wide)

car::vif(lm_vif)
```

---

## 5. GLOBAL GAM MODELS (one per response variable)

### 5.1 Fit one model per response (LCZ included as a control covariate)

```{r global-gam-fit}
fit_global_gam <- function(response_var, data) {
  formula_str <- paste0(
    response_var, " ~ s(Tree_Canopy) + s(Impervious) + s(Bldg_Footprint) + LCZ"
  )

  model_data <- data %>% drop_na(all_of(response_var))

  gam(as.formula(formula_str), data = model_data, method = "REML")
}

global_models <- map(RESPONSE_VARS, ~fit_global_gam(.x, lst_predictors_complete))
names(global_models) <- RESPONSE_VARS
```

### 5.2 Model summary comparison table (side-by-side across responses)

```{r global-gam-summary-table}
global_summary <- map_dfr(RESPONSE_VARS, function(resp) {
  s <- summary(global_models[[resp]])
  tibble(
    Response = resp,
    R2_adj = s$r.sq,
    Deviance_explained = s$dev.expl,
    n_obs = nrow(global_models[[resp]]$model)
  )
})

print(global_summary)
```

### 5.3 Smooth-term significance comparison (wide table: term x response)

```{r global-gam-term-significance}
extract_smooth_terms <- function(model, response_name) {
  s <- summary(model)
  as.data.frame(s$s.table) %>%
    rownames_to_column("Term") %>%
    mutate(Response = response_name) %>%
    select(Term, Response, edf, `p-value`)
}

global_term_long <- map2_dfr(global_models, RESPONSE_VARS, extract_smooth_terms)

global_pvalue_wide <- global_term_long %>%
  select(Term, Response, `p-value`) %>%
  pivot_wider(names_from = Response, values_from = `p-value`)

global_edf_wide <- global_term_long %>%
  select(Term, Response, edf) %>%
  pivot_wider(names_from = Response, values_from = edf)

cat("=== Smooth term p-values (by response) ===\n")
print(global_pvalue_wide)

cat("\n=== Smooth term EDF (by response) ===\n")
print(global_edf_wide)
```

### 5.4 Concurvity check (using the Average model as representative)

```{r global-concurvity}
concurvity(global_models[["Average"]], full = TRUE)
```

### 5.5 Diagnostics for each response model

```{r global-diagnostics}
for (resp in RESPONSE_VARS) {
  cat("\n=== gam.check:", resp, "===\n")
  gam.check(global_models[[resp]])
}
```

### 5.6 Visualize smooth effects — grid arranged for direct comparison

```{r global-smooth-plots, fig.width=12, fig.height=14}
par(mfrow = c(length(RESPONSE_VARS), 3))
for (resp in RESPONSE_VARS) {
  plot(global_models[[resp]], select = 1, shade = TRUE, seWithMean = TRUE,
       main = paste(resp, "- Tree Canopy"))
  plot(global_models[[resp]], select = 2, shade = TRUE, seWithMean = TRUE,
       main = paste(resp, "- Impervious"))
  plot(global_models[[resp]], select = 3, shade = TRUE, seWithMean = TRUE,
       main = paste(resp, "- Bldg Footprint"))
}
```

### 5.7 Predictor importance comparison (global models)

```{r global-importance}
extract_term_importance <- function(model, response_name,
                                     predictor_labels = c("Tree_Canopy", "Impervious", "Bldg_Footprint")) {
  terms_mat <- predict(model, type = "terms")
  term_names <- colnames(terms_mat)

  map_dfr(predictor_labels, function(pred) {
    match_name <- term_names[grepl(pred, term_names, fixed = TRUE)]
    if (length(match_name) == 0) {
      return(tibble(Response = response_name, Predictor = pred,
                     Effect_Range = NA_real_, Effect_SD = NA_real_))
    }
    vals <- terms_mat[, match_name[1]]
    tibble(
      Response = response_name,
      Predictor = pred,
      Effect_Range = max(vals) - min(vals),
      Effect_SD = sd(vals)
    )
  })
}

global_importance <- map2_dfr(global_models, RESPONSE_VARS, extract_term_importance)

# relative importance (%) within each response — which predictor dominates
global_importance <- global_importance %>%
  group_by(Response) %>%
  mutate(Relative_Importance_Pct = 100 * Effect_Range / sum(Effect_Range)) %>%
  ungroup()

global_importance_wide <- global_importance %>%
  select(Predictor, Response, Effect_Range) %>%
  pivot_wider(names_from = Response, values_from = Effect_Range)

cat("=== Effect range (°C) by predictor and response — directly comparable across responses ===\n")
print(global_importance_wide)

cat("\n=== Relative importance (% of total effect range within each response) ===\n")
print(global_importance %>% select(Response, Predictor, Relative_Importance_Pct) %>%
        pivot_wider(names_from = Response, values_from = Relative_Importance_Pct))
```

```{r global-importance-plot, fig.width=10, fig.height=5}
ggplot(global_importance, aes(x = Response, y = Effect_Range, fill = Predictor)) +
  geom_col(position = "dodge") +
  labs(
    title = "Predictor Effect Range by Response Variable (Global Models)",
    subtitle = "Larger range = greater influence on LST across its observed range",
    y = "Effect Range (°C)",
    x = "Response"
  ) +
  theme_minimal()
```

---

## 6. LCZ-STRATIFIED GAM MODELS (four responses x each LCZ class)

### 6.1 Identify LCZ classes with sufficient data

```{r lcz-adequate-classes}
min_obs <- 100

lcz_counts <- lst_predictors_complete %>%
  st_drop_geometry() %>%
  count(LCZ) %>%
  arrange(desc(n))

print(lcz_counts)

lcz_adequate <- lcz_counts %>%
  filter(n >= min_obs) %>%
  pull(LCZ)

cat("\n=== LCZ classes with >=", min_obs, "observations ===\n")
print(lcz_adequate)
```

### 6.2 Fit one model per (response x LCZ) combination

```{r lcz-gam-fit}
fit_lcz_gam <- function(response_var, lcz_class, data) {
  lcz_data <- data %>%
    filter(LCZ == lcz_class) %>%
    drop_na(all_of(response_var))

  formula_full <- as.formula(paste0(
    response_var, " ~ s(Tree_Canopy, k = 5) + s(Impervious, k = 5) + s(Bldg_Footprint, k = 5)"
  ))
  formula_reduced <- as.formula(paste0(
    response_var, " ~ s(Tree_Canopy, k = 4) + s(Impervious, k = 4)"
  ))

  model <- tryCatch({
    gam(formula_full, data = lcz_data, method = "REML")
  }, error = function(e) {
    gam(formula_reduced, data = lcz_data, method = "REML")
  })

  return(model)
}

# build a (response x LCZ) grid of models
lcz_model_grid <- expand_grid(Response = RESPONSE_VARS, LCZ = lcz_adequate) %>%
  mutate(model = map2(Response, LCZ, ~fit_lcz_gam(.x, .y, lst_predictors_complete)))
```

### 6.3 Model performance comparison table (LCZ x response, side-by-side)

```{r lcz-performance-table}
lcz_performance <- lcz_model_grid %>%
  mutate(
    R2_adj = map_dbl(model, ~summary(.x)$r.sq),
    Dev_expl = map_dbl(model, ~summary(.x)$dev.expl),
    n_obs = map_dbl(model, ~nrow(.x$model))
  ) %>%
  select(Response, LCZ, R2_adj, Dev_expl, n_obs)

r2_wide <- lcz_performance %>%
  select(LCZ, Response, R2_adj) %>%
  pivot_wider(names_from = Response, values_from = R2_adj)

dev_wide <- lcz_performance %>%
  select(LCZ, Response, Dev_expl) %>%
  pivot_wider(names_from = Response, values_from = Dev_expl)

cat("=== Adjusted R^2 by LCZ and response ===\n")
print(r2_wide)

cat("\n=== Deviance explained by LCZ and response ===\n")
print(dev_wide)
```

### 6.4 Predictor importance comparison (LCZ x response x predictor)

```{r lcz-importance-table}
lcz_importance <- lcz_model_grid %>%
  mutate(importance = map2(model, Response, ~extract_term_importance(.x, .y) %>% select(-Response))) %>%
  unnest(importance) %>%
  select(LCZ, Response, Predictor, Effect_Range, Effect_SD) %>%
  group_by(LCZ, Response) %>%
  mutate(Relative_Importance_Pct = 100 * Effect_Range / sum(Effect_Range, na.rm = TRUE)) %>%
  ungroup()

cat("=== Full LCZ x Response x Predictor importance table ===\n")
print(lcz_importance)
```

### 6.5 Which predictor dominates in each LCZ class? (summary across responses)

```{r lcz-importance-dominant-predictor}
lcz_dominant_predictor <- lcz_importance %>%
  group_by(LCZ, Response) %>%
  slice_max(Effect_Range, n = 1) %>%
  ungroup() %>%
  select(LCZ, Response, Dominant_Predictor = Predictor, Effect_Range)

dominant_wide <- lcz_dominant_predictor %>%
  select(LCZ, Response, Dominant_Predictor) %>%
  pivot_wider(names_from = Response, values_from = Dominant_Predictor)

cat("=== Dominant predictor by LCZ and response ===\n")
print(dominant_wide)
```

### 6.6 Visualize importance — heatmap style, faceted by response

```{r lcz-importance-heatmap, fig.width=12, fig.height=8}
ggplot(lcz_importance, aes(x = LCZ, y = Predictor, fill = Effect_Range)) +
  geom_tile(color = "white") +
  facet_wrap(~ Response, ncol = 2) +
  scale_fill_gradient(low = "white", high = "firebrick", name = "Effect Range (°C)") +
  labs(
    title = "Predictor Effect Range by LCZ Class and Response Variable",
    x = "LCZ Class", y = "Predictor"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 6.7 Diagnostics for each LCZ-stratified model

```{r lcz-diagnostics}
walk2(lcz_model_grid$model, paste(lcz_model_grid$Response, lcz_model_grid$LCZ, sep = " / LCZ "),
      function(model, label) {
        cat("\n=== gam.check:", label, "===\n")
        gam.check(model)
      })
```

---

## 7. COMPARATIVE SUMMARY: Global vs LCZ-Stratified

### 7.1 Performance comparison, all responses side-by-side

```{r comparative-performance-summary}
global_perf_long <- global_summary %>%
  mutate(Model_Type = "Global") %>%
  select(Response, Model_Type, R2_adj, Deviance_explained)

lcz_perf_avg <- lcz_performance %>%
  group_by(Response) %>%
  summarise(
    R2_adj = mean(R2_adj, na.rm = TRUE),
    Deviance_explained = mean(Dev_expl, na.rm = TRUE)
  ) %>%
  mutate(Model_Type = "LCZ-Stratified (avg across classes)")

comparative_summary <- bind_rows(global_perf_long, lcz_perf_avg) %>%
  select(Model_Type, Response, R2_adj, Deviance_explained) %>%
  arrange(Response, Model_Type)

print(comparative_summary)
```

```{r comparative-performance-plot, fig.width=10, fig.height=5}
ggplot(comparative_summary, aes(x = Response, y = R2_adj, fill = Model_Type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Global vs LCZ-Stratified Model Fit, by Response Variable",
    y = "Adjusted R-squared", x = "Response"
  ) +
  theme_minimal()
```

### 7.2 Which predictor matters most overall? (global importance summary)

```{r comparative-importance-summary}
overall_importance_summary <- global_importance %>%
  select(Response, Predictor, Effect_Range, Relative_Importance_Pct) %>%
  arrange(Response, desc(Effect_Range))

print(overall_importance_summary)

cat("\n=== Average relative importance across all four responses ===\n")
overall_importance_summary %>%
  group_by(Predictor) %>%
  summarise(Avg_Relative_Importance_Pct = mean(Relative_Importance_Pct)) %>%
  arrange(desc(Avg_Relative_Importance_Pct)) %>%
  print()
```

---

## 8. EXPORT RESULTS

```{r export-results}
if (!dir.exists("output")) {
  dir.create("output")
}

write.csv(global_summary, "output/global_model_summary.csv", row.names = FALSE)
write.csv(global_pvalue_wide, "output/global_smooth_pvalues.csv", row.names = FALSE)
write.csv(global_importance_wide, "output/global_predictor_effect_range.csv", row.names = FALSE)
write.csv(lcz_performance, "output/lcz_model_performance.csv", row.names = FALSE)
write.csv(lcz_importance, "output/lcz_predictor_importance.csv", row.names = FALSE)
write.csv(dominant_wide, "output/lcz_dominant_predictor.csv", row.names = FALSE)
write.csv(comparative_summary, "output/comparative_summary.csv", row.names = FALSE)

cat("\nResults exported to output/ directory\n")
```